Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
some changes to the related work

In addition to working with effects, compiler intermediate representations also
have to track accesses to memory. This information is typically gathered via
ad-hoc alias analyses, but a compositional alternative to these analyses is to
track this information via substructural type systems such as linear and affine
types, which associate aliasing and ownership information with individual
substructural values. Having aliasing information justifies important
optimisations such as loop hoisting: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
while (e) {
  x = read(a);
  body
}
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
x = read(a);
while(e) {
  body
}
\end{BVerbatim}
\label{eqn:hoist}
\end{equation}
This transformation is only sound if the body of the loop \texttt{body} does not
access \texttt{a}, which is precisely the kind of frame condition that linear
types can track. Moreover, since we already want to track whether effectful
\emph{expressions} can be duplicated or discarded, the very same machinery can
be reused to track whether \emph{values} are linear/affine/relevant.

Historically, compiler writers have treated the semantics of their IR quite
informally, since the "intended model" of the language is a simple sequential
machine operating over a heap which is a flat array of bytes. Unfortunately,
over time both the intended model and the actual hardware have drifted away from
this simple mental model. Compiler frameworks have to model concurrent accesses
to memory in order to compile multithreaded code, and furthermore the IR's
memory model has to be sound with respect to the actual weak memory models
supported by hardware.

@@

Our main metatheoretic results will be \textit{substitution} and
\textit{label-substitution}, which will require introducing some new
metatheoretic judgements to state properly, which we do in Figure
\ref{fig:ssa-meta-judgements}. In particular, the judgement
\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is taken to be
a map from variable names to expressions, is a valid substitution taking
\(\Theta\) to \(\Gamma\). What this means is that, for each variable
\(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can be interpreted as a
\textit{central} term of type \(A^q\) in a subcontext of \(\Theta\), such that
when this is done for all \(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables
in \(\Theta\) are used up: this is described formally via the typing rules
\rle{subst-nil}, which says that any function \(\gamma\) takes the empty context
to itself, and \rle{subst-cons}, which says that if \(\gamma\) takes
\(\Theta_\Gamma\) to \(\Gamma\), \(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\), \(q
\subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the duplication/discarding
of substructural variables), and \(\Theta\) splits into \(\Theta_\Gamma\) and
\(\Theta_x\), then \(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).

@@

\section{Related Work}

\TODO{\cite{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}

\TODO{\cite{schneider-phd}}

\TODO{\cite{chakravarty-functional-ssa-2003}}

\TODO{\cite{barthe-compcert-ssa-2014}}

\TODO{\cite{buchwald-verified-ssa-2016}}

\TODO{\cite{herklotz-gsa-2023}}

\TODO{\cite{pop-ssa-inout-2009}}

\TODO{\cite{garbuzov-structural-cfg-2018}}

\TODO{\cite{hua-explicit-ssa-2010}}

\TODO{\cite{rvsdg}}

>>>>

In addition to working with effects, compiler intermediate representations also
have to track accesses to memory. This information is typically gathered via
ad-hoc alias analyses, but a compositional alternative to these analyses is to
track this information via substructural type systems such as linear and affine
types, which associate aliasing and ownership information with individual
substructural values. Having aliasing information justifies important
optimisations such as loop hoisting: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
while (e) {
  x = load(a);
  body
}
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
x = load(a);
while(e) {
  body
}
\end{BVerbatim}
\label{eqn:hoist}
\end{equation}
This transformation is only sound if the body of the loop \texttt{body} does not
access \texttt{a}, which is precisely the kind of frame condition that linear
types can track. Moreover, since we already want to track whether effectful
\emph{expressions} can be duplicated or discarded, the very same machinery can
be reused to track whether \emph{values} are linear/affine/relevant.

Historically, compiler writers have treated the semantics of their IR quite
informally, since the "intended model" of the language is a simple sequential
machine operating over a flat array of bytes. Unfortunately,
over time both the intended model and the actual hardware have drifted away from
this simple mental model. Compiler frameworks have to model concurrent accesses
to memory in order to compile multithreaded code, and furthermore the IR's
memory model has to be sound with respect to the actual weak memory models
supported by hardware.

@@

Our main metatheoretic results will be \textit{substitution} and
\textit{label-substitution}, which will require introducing some
substitution judgements to state properly, which we do in Figure
\ref{fig:ssa-meta-judgements}. In particular, the judgement
\(\issubst{\gamma}{\Theta}{\Gamma}\) says that \(\gamma\), which is
taken to be a map from variable names to expressions, is a valid
substitution taking \(\Theta\) to \(\Gamma\). What this means is that,
for each variable \(\thyp{x}{A}{q}\) in \(\Gamma\), \(\gamma(x)\) can
be interpreted as a \textit{central} term of type \(A^q\) in a
subcontext of \(\Theta\), such that when this is done for all
\(\thyp{x}{A}{q} \in \Gamma\) any non-affine variables in \(\Theta\)
are used up: this is described formally via the typing rules
\rle{subst-nil}, which says that any function \(\gamma\) takes the
empty context to itself, and \rle{subst-cons}, which says that if
\(\gamma\) takes \(\Theta_\Gamma\) to \(\Gamma\),
\(\hasty{\Theta_x}{1}{\gamma(x)}{A}{q}\),
\(q \subseteq \ms{lin}(\Theta_x)\) (to avoid necessating the
duplication/discarding of substructural variables), and \(\Theta\)
splits into \(\Theta_\Gamma\) and \(\Theta_x\), then
\(\issubst{\gamma}{\Theta}{\Gamma, \thyp{x}{A}{q}}\).

@@

\section{Related Work and Discussion}

There is a vast literature on SSA and related compiler IRs, but for
space reasons, we are forced to restrict our discussion to the most
closely related work.

\paragraph{Weak memory}

\TODO{\citet{leaky-semicolon} for weak memory stuff: future work: add while loops, and also refinement}
\TODO{\citet{paviotti-modular-relaxed-dep-20}}

\todo{Main thing here to say is that we need semantics for while loops, plus associativity of semicolon, if statements must support distributivity}

\paragraph{Type Theory for SSA}

\TODO{\citet{hua-explicit-ssa-2010}}
\TODO{\citet{ssa-types-matsuno-ohori-06}}
\TODO{\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20}}

\paragraph{Functional SSA}

\TODO{\citet{chakravarty-functional-ssa-2003}}

\subparagraph{Operational Semantics}

\TODO{\citet{barthe-compcert-ssa-2014}}
\TODO{\citet{buchwald-verified-ssa-2016}}
\TODO{\citet{herklotz-gsa-2023}}
\TODO{\citet{garbuzov-structural-cfg-2018}}
\TODO{\citet{vellum}}

\subparagraph{Denotational Semantics}

\TODO{\citet{pop-ssa-inout-2009}}

% \TODO{\cite{schneider-phd}}
% \TODO{\cite{rvsdg}}
